A Plasma Torus Around a Young Low-Mass Star

### 2025-03-24 paper/ms.tex
update

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
coherent sinusoidal Balmer emission at twice to four times the
star's equatorial velocity, demonstrating the existence of
corotating clumps of circumstellar plasma --- a plasma torus.  Given
that long-lived condensations of cool ($10^4$ K) plasma can persist
in the hot ($10^6$ K) coronae of stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can sustain an optically thick component; the origin
of the optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Community investment with JWST has proceeded accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has in turn become a major theme in
exoplanet and stellar astrophysics.  Previous work has established
that most M dwarfs host close-in planets \cite{Dressing2015} that are
on average subject to long circumstellar disk lifetimes
\cite{Ribas2015}, to large doses of UV radiation \cite{France2013},
and to a high incidence of flares and coronal mass ejections
\cite{Gunther2020}.  However, despite the excellent work in these areas,
the properties of the plasma and magnetospheric environments that
bathe young, close-in exoplanets have remained challenging to
quantify.

@@

% * keck/hires: behavior at v/veq<1
The stellar H$\alpha$ line core, at $|\Delta v / v_{\rm
eq}|<1$, captures a superposition of emission from the star's
chromosphere, modulated by any occulting material capable of absorbing
or emitting H$\alpha$ photons.  In Figure~\ref{fig:spec}e, the data
from $\phi$=0.4-1.2 have a simple interpretation: from $\phi$=0.4-0.9,
a hot (emitting) region first gradually crosses the stellar line
profile, followed from $\phi$=0.7-1.2 by the transit of a cool
(absorbing) region. Phases $\phi$$<$0.4 show a mix of similar events,
though the time sampling is sufficiently coarse that the
interpretation is less clear.  A final exercise to quantify the
behavior in the line core is shown in Figure~\ref{fig:spec}d, where
$f_{\rm H\alpha\ core}$ denotes the summed flux at $|\Delta v / v_{\rm
eq}|<1$.  This panel shows that changes in the line core flux are
correlated with the broadband variability over most of the light
curve, except at $\phi$$\approx$0.5, during the transit of the
3.9\,$R_\star$ clump and the occultation of the lower-velocity clump.

@@

Spectra of magetically-active, rapidly rotating stars with a wide range
of masses have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} and
transient absorption features
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for
such spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar corona
that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the Sun's surface because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate over much
longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit, nor be optically thick.
For instance, the H$\alpha$ emission clump at 3.9\,$R_\star$ shows no 
photometric signal in the February 2024 photometry.

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they show that corotating circumstellar plasma clumps
are present around at least one such star.  This observation rules out
a ``starspot-only'' origin scenario for CPVs, \cite{Koen2021} because
such scenarios have no means of explaining spectroscopic emission
beyond the stellar disk.  Similarly, scenarios in which the
circumstellar material is made only of dust are also ruled out.  While
dust may be present, to explain the H$\alpha$ emission the clumps must
include plasma with a significant population of hydrogen atoms in the
$n$=3 excited state, for $n$ the principal quantum number.  More quantitatively, the observed H$\alpha$
luminosity suggests characteristic densities and masses for the
gaseous component of these clumps of $n_{\rm H} \sim
10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm gas} \sim 10^{17}$\,g (see
Section~\ref{subsec:gas}).  Dust, if present, is independently
constrained to have a total mass $10^{15}\,{\rm g} < M_{\rm dust} <
10^{17}\,{\rm g}$ (see Section~\ref{subsec:dust}).  This circumstellar
material could originate either from the star or from an external
source.  Plausible external sources include an old and undetected
disk, comets, or a close-in exoplanet.
This latter possibility would render CPVs as extrasolar analogs of the
Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

@@

The role of the star's magnetic field could be better understood
through new observations, and new theory.  From the theoretical
perspective, there is an urgent need for rigid-field
magneto-hydrodynamic modeling to go beyond previous work
\cite{Townsend2005,Townsend2008,Krticka2022} and to explore what field
topologies might explain the observed light curve morphologies
\cite{Bouma2024}.  In particular, dynamo simulations of
fully-convective M dwarfs have suggested that global-scale mean fields
might be confined to a single hemisphere \cite{Brown2020}; such fields
would yield accumulation surfaces quite different from those that have
previously been explored.  Observationally, spectropolarimetry has the
potential to assess both the field strength and topology.  An
independent probe could also be connected the recent work
\cite{Kaur2024} showing that CPVs are variable radio emitters with
emission components that can be not only persistent, but also
short-lived and highly polarized.  In particular, detecting radio
emission produced by an electron cyclotron maser instability (e.g.
\cite{Callingham2021}) would provide a measurement of the field
strength at the site of the emitting region, further clarifying the
magnetic field's structure.

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs
\cite{Dressing2015}.  However, 0.3-3\% of young M dwarfs show the CPV
phenomenon \cite{Rebull2020}, and our data show that the complex
photometric features occur when clumps of circumstellar material
transit the star.  The implied geometric correction suggests that an
appreciable minority (10-30\%) of young M dwarfs -- the rapidly
rotating ones with centrifugal magnetospheres -- host similar
circumstellar environments to the CPVs.

@@

{\it Spectroscopy:}
We observed TIC~141146667 ($V$=16.2) with Keck/HIRES for five hours over
a second-half night spanning 17 February 2024 10:47 to 16:13 (UT).
The star's airmass over this window spanned $z$=1.2-2.2, and we opted
for a fixed 15 minute cadence over the entire sequence, except for a
final 10 minute exposure due to increasing sky brightness at sunrise.
We observed without the iodine cell and used the C2 decker
(0$\farcs$86$\times$14$\farcs$0) in the red instrument configuration,
yielding a spectral resolution $R$$\approx$45{,}0000 ($\delta
v$$\approx$6.7\,\kms; $\delta v / v_{\rm eq}$$\approx$0.05).  We binned
the CCD readout by a factor of three in the spatial dimension, yielding
$\approx$1,000 photons (S/N=33) per pixel in the continuum at 6500\,\AA,
at minimum airmass.  Strong winds contributed to 1\farcs2$\pm$0\farcs2
seeing over the night, but conditions were otherwise favorable.  We
reduced the echelleogram to a one-dimensional spectrum using the
standard techniques of the California Planet Survey \cite{Howard2010}.
Figure~\ref{fig:spec}b shows the result in the vicinity of H$\alpha$
without any additional processing.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=\textwidth]{figures/sf5.pdf}
  \caption{Time evolution of select regions in the Keck/HIRES
  spectra from 17 February 2025.
  We made this plot by applying a windowed outlier rejection to remove
  cosmic rays and then smoothing each spectrum with a Gaussian filter.
  The horizontal axis shows the velocity relative to each
  line's rest wavelength, normalized by the stellar equatorial
  velocity.  
  H$\alpha$ is the only Balmer line to show periodic variability of
  the form seen in Figure~\ref{fig:spec}.
  The He 5876\,\AA\ line shows a time-dynamic blueshift that 
  differs from the Balmer lines.
  Li 6708\,\AA\ shows no obvious absorption.
  }
  \label{fig:hirescuts}
\end{figure}

Figure~\ref{fig:hirescuts} shows a few regions of interest in the
HIRES spectra, which cover 3650-7960\,\AA.  Higher order Balmer lines
including H$\gamma$ and H$\delta$ ($n=5\rightarrow2$ and
$n=6\rightarrow2$) do show some variability
outside the line core.  However, this variability is not clearly
periodic in the same manner as the emission seen in H$\alpha$.  This
could be because there are insufficient hydrogen ions recombining from
the relevant excited states, or due to the lower precision of the
spectra in these bluer regions.  The Ca HK doublet is
also detected in emission, however the continuum near it is not.
Chromospheric emission from the magnesium triplet are also
detected, but are sufficiently blended to be rendered unusable.

A sudden blue excess appeared at $\phi$$\approx$0.2 in H$\alpha$,
H$\gamma$, H$\delta$, and He 5876\,\AA.  At the same epoch, only
H$\alpha$ clearly shows a resolved red excess as well, although
H$\gamma$ and H$\delta$ do broaden at this time.  Given the sudden
rise of this emission in $<$15\,minutes, and the association between
He 5876\,\AA\ and stellar outflows at high excitation energy, this
component of the spectra might be associated with a flare, rather
than with a stable, periodic component.

@@

The physical conditions inside the plasma clumps, in particular the
hydrogen number density, plasma temperature, ionization fraction, and
magnetic field strength can be estimated from the available data.
We caution that the following estimates are simple order of magnitude
calculations: detailed considerations of radiative transfer are a
worthy topic for future work, but are beyond the scope of this
article.

>>>>

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
coherent sinusoidal Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time the existence of corotating circumstellar plasma around a CPV.
Given that long-lived condensations of cool ($10^4$ K) plasma can
persist in the hot ($10^6$ K) coronae of stars with a wide range of
masses \cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can sustain an optically thick component; the origin
of the optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Community investment with JWST has proceeded accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has in turn become a major theme in
exoplanet and stellar astrophysics.  Previous work has established
that most M dwarfs host close-in planets \cite{Dressing2015} that are
on average subject to long circumstellar disk lifetimes
\cite{Ribas2015}, to large doses of UV radiation \cite{France2013},
and to a high incidence of flares and coronal mass ejections
\cite{Gunther2020}.  However, despite extensive work in these areas,
the plasma and magnetospheric environments that bathe young, close-in
exoplanets remain challenging to quantify.
Understanding these environments is crucial because they directly
impact atmospheric retention and habitability of close-in planets.

@@

% * keck/hires: behavior at v/veq<1
The stellar H$\alpha$ line core, at $|\Delta v / v_{\rm
eq}|<1$, captures a superposition of emission from the star's
chromosphere, modulated by any occulting material capable of absorbing
or emitting H$\alpha$ photons.  In Figure~\ref{fig:spec}e, the data
from $\phi$=0.4-1.2 have a simple interpretation: from $\phi$=0.4-0.9,
a hot (emitting) region first gradually crosses the stellar line
profile, followed from $\phi$=0.7-1.2 by the transit of a cool
(absorbing) region. Phases $\phi$$<$0.4 show a mix of similar events,
though the time sampling is sufficiently coarse that the
interpretation is less clear.  A final exercise to quantify the
behavior in the line core is shown in Figure~\ref{fig:spec}d, where
$f_{\rm H\alpha\ core}$ denotes the summed flux at $|\Delta v / v_{\rm
eq}|<1$.  This panel shows that changes in the line core flux 
correlate with the broadband variability throughout most of the light
curve, except near $\phi$$\approx$0.5, corresponding to the transit of the
3.9\,$R_\star$ clump and the occultation of the lower-velocity clump.

@@

Spectra of magnetically-active, rapidly rotating stars with a wide range
of masses have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} and
transient absorption features
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for
such spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar corona
that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the Sun's surface because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate over much
longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit, nor be optically thick.
For instance, the H$\alpha$ emission clump at 3.9\,$R_\star$ shows no 
photometric signal in the February 2024 photometry.

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they show that corotating circumstellar plasma clumps
are present around at least one such star.  This observation rules out
a ``starspot-only'' origin scenario for CPVs, \cite{Koen2021} because
such scenarios have no means of explaining spectroscopic emission
beyond the stellar disk.  Similarly, scenarios in which the
circumstellar material is made only of dust are also ruled out.  While
dust may be present, to explain the H$\alpha$ emission the clumps must
include plasma with a significant population of hydrogen atoms in the
$n$=3 excited state, for $n$ the principal quantum number.  More quantitatively, the observed H$\alpha$
luminosity suggests characteristic densities and masses for the
gaseous component of these clumps of $n_{\rm H} \sim
10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm gas} \sim 10^{17}$\,g (see
Section~\ref{subsec:gas}).  Dust, if present, is independently
constrained to have a total mass $10^{15}\,{\rm g} < M_{\rm dust} <
10^{17}\,{\rm g}$ (see Section~\ref{subsec:dust}).  This circumstellar
material could originate either from the star or from an external
source.  Plausible external sources include an old and undetected
disk, comets, or a close-in exoplanet.
This latter scenario would make CPVs extrasolar analogs of the
Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

@@

The role of the star's magnetic field could be better understood
through new observations, and new theory.  From the theoretical
perspective, there is an urgent need for rigid-field
magnetohydrodynamic modeling to go beyond previous work
\cite{Townsend2005,Townsend2008,Krticka2022} and to explore what field
topologies might explain the observed light curve morphologies
\cite{Bouma2024}.  In particular, dynamo simulations of
fully-convective M dwarfs have suggested that global-scale mean fields
might be confined to a single hemisphere \cite{Brown2020}; such fields
would yield accumulation surfaces quite different from those that have
previously been explored.  Observationally, spectropolarimetry has the
potential to assess both the field strength and topology.  An
independent constraint on magnetic field topology may also come from
radio observations, which have shown \cite{Kaur2024} that CPVs emit
variable radio signals, including persistent and polarized components.
Detecting radio emission produced by an electron cyclotron maser
instability (e.g. \cite{Callingham2021}) in particular would provide
a measurement of the field strength at the site of the emitting
region, further clarifying the magnetosphere's structure.

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs
\cite{Dressing2015}.  However, 0.3-3\% of young M dwarfs show the CPV
phenomenon \cite{Rebull2020}, and our data show that the complex
photometric features occur when clumps of circumstellar material
transit the star.  The implied geometric correction suggests that an
appreciable minority (10-30\%) of young M dwarfs -- the rapidly
rotating ones with centrifugal magnetospheres -- host similar
circumstellar environments to the CPVs.
Future studies that combine spectroscopic, polarimetric, and
multi-wavelength observations, along with magnetohydrodynamic
modeling, will be key to understanding the complex magnetospheric
environments of these young stars.

@@

{\it Spectroscopy:}
We observed TIC~141146667 ($V$=16.2) with Keck/HIRES for five hours
during the second half of the night, spanning 17 February 2024 10:47
to 16:13 (UT).
The star's airmass over this window spanned $z$=1.2-2.2, and we opted
for a fixed 15 minute cadence over the entire sequence, except for a
final 10 minute exposure due to increasing sky brightness at sunrise.
We observed without the iodine cell and used the C2 decker
(0$\farcs$86$\times$14$\farcs$0) in the red instrument configuration,
yielding a spectral resolution $R$$\approx$45{,}0000 ($\delta
v$$\approx$6.7\,\kms; $\delta v / v_{\rm eq}$$\approx$0.05).  We binned
the CCD readout by a factor of three in the spatial dimension, yielding
$\approx$1,000 photons (S/N=33) per pixel in the continuum at 6500\,\AA,
at minimum airmass.  Strong winds contributed to 1\farcs2$\pm$0\farcs2
seeing over the night, but conditions were otherwise favorable.  We
reduced the echelleogram to a one-dimensional spectrum using the
standard techniques of the California Planet Survey \cite{Howard2010}.
Figure~\ref{fig:spec}b shows the result in the vicinity of H$\alpha$
without any additional processing.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=\textwidth]{figures/sf5.pdf}
  \caption{Time evolution of select regions in the Keck/HIRES
  spectra from 17 February 2024.
  We made this plot by applying a windowed outlier rejection to remove
  cosmic rays and then smoothing each spectrum with a Gaussian filter.
  The horizontal axis shows the velocity relative to each
  line's rest wavelength, normalized by the stellar equatorial
  velocity.  
  H$\alpha$ is the only Balmer line to show periodic variability of
  the form seen in Figure~\ref{fig:spec}.
  The He 5876\,\AA\ line shows a time-dynamic blueshift that 
  differs from the Balmer lines.
  Li 6708\,\AA\ shows no obvious absorption.
  }
  \label{fig:hirescuts}
\end{figure}

Figure~\ref{fig:hirescuts} shows a few regions of interest in the
HIRES spectra, which cover 3650-7960\,\AA.  Higher order Balmer lines
including H$\gamma$ and H$\delta$ ($n=5\rightarrow2$ and
$n=6\rightarrow2$) do show variability outside the line core.
However, this variability is not clearly periodic in the same manner
as the emission seen in H$\alpha$.  This could be because there are
insufficient hydrogen ions in the relevant excited states, or because
the spectra have lower precision in these bluer regions.  The Ca HK
doublet is also detected in emission, while the continuum near it is
not.  Chromospheric emission from the magnesium triplet is also
detected, but these lines are sufficiently blended to be rendered
unusable.

Figure~\ref{fig:hirescuts} provides a novel view on the sudden blue
excess that appeared at $\phi$$\approx$0.2 in H$\alpha$, H$\gamma$,
H$\delta$, and He 5876\,\AA.  At the same epoch, only H$\alpha$
clearly shows a resolved red excess as well, although H$\gamma$ and
H$\delta$ are broadened at this time.  Given the sudden rise of this
emission in $<$15\,minutes, this sudden blue excess might suggest a
more sudden flow, rather than a stable, periodic component.  A stellar
flare could be a plausible explanation for such a sudden rise.
However this idea seems incompatible with the sinusoidal emission seen
from $\phi$$\approx$0.5-1.0, and with the fact that flares often
excite iron lines in the blue HIRES orders, which are not observed.
More time-series spectroscopy would be needed to provide a definitive
clarification for this type of non-periodic variability.

@@

The physical conditions inside the plasma clumps, in particular the
hydrogen number density, plasma temperature, ionization fraction, and
magnetic field strength can be estimated from the available data.
We caution that the following estimates are order of magnitude
calculations that assume a simple uniform-density plasma: detailed
considerations of radiative transfer are a worthy topic for future
work, but are beyond the scope of this article.
